---
title: "texas-extreme-weather"
format: html
message: false
warning: false
---

```{r}
#| output: false

# Load in libraries
library(tidyverse)
library(tmap) # map making
library(sf)
library(stars)
library(terra) # raster handling
```
Reading in Nasa's Worldview data that will be used to analyze the effects of the power outages in Texas. February 7, 2021 and February 16, 2021 are the two dates analyzed as they provide clear skies for better analysis.
```{r}
# Read in the NOAA data
nasa1 <- read_stars(here::here("data/VNP46A1/VNP46A1.A2021038.h08v05.001.2021039064328.tif"))
nasa2 <- read_stars(here::here("data/VNP46A1/VNP46A1.A2021038.h08v06.001.2021039064329.tif"))
nasa3 <- read_stars(here::here("data/VNP46A1/VNP46A1.A2021047.h08v05.001.2021048091106.tif"))
nasa4 <- read_stars(here::here("data/VNP46A1/VNP46A1.A2021047.h08v06.001.2021048091105.tif"))
```
Road data in Texas is necessary to take account of light pollution that comes from roads. 
```{r}
# Read in road data
roads <- read_sf(here::here("data/gis_osm_roads_free_1.gpkg"))

```
Part of this analysis will quantify the number of houses that experienced blackouts.
```{r}
# Read in house data
houses <- read_sf(here::here("data/gis_osm_buildings_a_free_1.gpkg"))
```


Initial exploration of the NASA Worldview data
```{r}
noaa_map1 <- tm_shape(nasa1) +
  tm_raster( )

noaa_map1
```

